plyranges: a grammar for transforming genomics data

### 2018-04-23 paper.Rmd
array example

# a BamFile from the H1 cell line
h1_bam <- "data/GSM433167_BI.H3K27me3.bam"

# Genotype Files accross h1 cell line methylation marks
h1_geno <- "data/GSM1463263_JS-ESCH1.txt.gz"

```
# Introduction

@@

@@

## Transition Transversion Ratio

(example of group_by + mutate + summarise and working on an object that inherits
from GRanges)

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed. This report was written with
`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio.

>>>>

# a BamFile from the H1 cell line
h1_bam <- "data/GSM433167_BI.H3K27me3.bam"
```
# Introduction

@@

## Quality Control Metrics

```{r, echo = FALSE, warning = FALSE, cache = TRUE}
# prepare a GRanges object from array data
library(readr)
# annotation information available here
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952
# chromosome zero are 'problem' probes and should be filtered
anno <- read_tsv("data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz",
                 trim_ws = TRUE,
                 col_types = c("cccc----")) %>%
  dplyr::filter(Chr != "0")

# genotypes for H1 cell line, data starts on line 10
# data obtained from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1463263
geno <- read_tsv("data/GSM1463263_JS-ESCH1.txt.gz", skip = 10)

# rename columns
geno <- geno %>% 
  dplyr::select(name = `SNP Name`, 
         score = `GC Score`, 
         baf = `B Allele Freq`, 
         logR = `Log R Ratio`)

# merge everything
complete_vranges_df <- dplyr::inner_join(geno, anno, by = c("name" = "Name")) %>%
  dplyr::rename(alleles = Alleles, seqnames = Chr, start = MapInfo)

h1_snp_array <- as_granges(complete_vranges_df,
                           start = as.integer(start),
                           end = as.integer(start))
h1_snp_array <- h1_snp_array %>% 
  mutate(ref = stringr::str_replace(alleles, "\\[(.)/.*", "\\1"),
         alt = stringr::str_replace(alleles, ".*/(.)\\]", "\\1"))
```

We have created a GRanges object from genotyping performed on the
H1 cell line, consisting of approximately two million single nucleotide
polymorphisms (SNP) and short insertion/deletions (indel). The GRanges object consists
of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
log relative intensity of the probes, GC content score over a probe, and the
name of the probe. We can use this information to compute the transition-transversion
ratio, a quality control metric, within each chromosome in GRanges object.

First we filter out any insertion or deletion alleles and 
the SNPs present on the mitochondria then create a logical vector corresponding
to whether there is a transition event. 
```{r}
h1_snp_array <- h1_snp_array %>%
  filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A")) |
                      (ref %in% c("C","T") & alt %in% c("T", "C")))
```

We can then compute the transition-transversion ratio using the `group_by`
and `summarise` pattern, it is computed as the total number of transition
SNPs divided by the total number of transversion SNPs within each chromsome
(figure \ref{fig:titv-viz}).

```{r}
ti_tv_results <- h1_snp_array %>% 
  group_by(seqnames) %>%
  summarise(n_snps = n(),
            ti_tv = sum(transition) / sum(!transition)) 
```

```{r titv-viz, echo = FALSE, fig.cap = "Dot plot of chromosomes ordered by estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome."}
ti_tv_results %>% 
  as.data.frame() %>% 
  ggplot(aes(forcats::fct_reorder(seqnames, ti_tv), ti_tv)) +
  geom_point() + 
  geom_hline(yintercept = 3, colour = "white", size = 4) +
  labs(x = "Chromosome", y = "Transition Transverion Ratio") +
  coord_flip()

@@

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed. This report was written with
`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio. All code
required to reproduce this report is available at 
https://github.com/sa-lee/plyranges-paper.
